---
title: "usrec_markdown"
author: "Jimmy Ieng"
date: "2025-10-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

#This is the Fredr R file converted to an R Markdown file to make the file more organized

```{r}

```

## installs the packages
```{r cars}
#install package, but rstudio can't locate it
#install.packages("fredr", dependencies = TRUE)
# make sure 00requirements.R is in the same folder your markdown file is in
# get the file of the requirements.R
source("00_requirements.R")
```

```{r}
#checks if R can find the fredr package
"fredr" %in% loadedNamespaces()
```


```{r cars}
# R can't find the fredr path, so we will have to manually find it

#this code checks where the path is located
# should be in a path called LibPath which starts with " and ends with "
installed.packages()["fredr",]
```

```{r cars}
# helps Rstudio finds the path
#replace current path with your path
.libPaths("/cloud/lib/x86_64-pc-linux-gnu-library/4.5")


#this is to check if its installed, if installed prints TRUE else FAlSE
"fredr" %in% loadedNamespaces()
```


#help menu for FREDR packages
```{r pressure, echo=FALSE}
?fredr_set_key
```

## unlocks access to Fred
```{r}
fredr_set_key("c859345561941fbef52080a4712c116b")
```


## Charles - "these are needed for every session"
```{r}
## these are needed for every session
#library(fredr)
#install.packages("fredr")
#library("ggplot2")
#install.packages("ggplot2")
#library(blsAPI)
```

## pulls up the fredr(years of UC recessions) data
```{r}
usrec <- fredr(series_id = "USREC")
head(usrec)
```

## filtered out data excluding the dates and recession indicator
```{r}
USrec_date <- usrec[,c("date","value")]
```




# pulls up the S&P 500 and the created GSPC_adj to get just the trading volume and the adjusted closing price
```{r}
#install.packages("quantmod")
#library(quantmod)
getSymbols("^GSPC") 

#Gets the index with the first day of the month
GSPC_monthly <- to.monthly(GSPC, indexAt = "lastof", OHLC = TRUE)
first_of_month <- as.Date(format(index(GSPC_monthly), "%Y-%m-01"))
index(GSPC_monthly) <- first_of_month

# Convert to data frame (keeps column names)
GSPC_monthly_df <- data.frame(date = index(GSPC_monthly), coredata(GSPC_monthly))

# Select only date, volume, and closing price
GSPC_adj <- GSPC_monthly_df[, c("date", "GSPC.Volume", "GSPC.Close")]

head(GSPC_adj)

```





#import unemployment_rates.csv from repository, then run this to rename the data

```{r}
names(unemployment_rates) <- c("date", "rate (%)")
```

#makes a new dataset with all the datasets combined
```{r}

combined <- Reduce(function(x, y) merge(x, y, by = "date", all = TRUE),
                   list(usrec, unemployment_rates, GSPC_adj))

combined <- combined[,-2]
names(combined)[c(2,5)] <- c("Recess", "UnempRate")
```

# gets the rows that has all the columns filled

# THIS WILL BE THE DATA WE WILL LOOK AT
```{r}
recession_indicators <- combined[complete.cases(combined),]
```

