---
title: "Final report"
author: "Guanlin Huang"
date: "2025-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Title and Abstract
Give a title to your project. Then, in an abstract of 200 words or less, describe what is the main question you aimed to answer, and a brief summary of your conclusions.

Answer Here:
Title: The Impact of the US Economic and Financial Environment

Abstract: This project aims to answer the following core question: By analyzing key US financial data (including GDP growth rate, inflation rate, unemployment rate, interest rates, and stock market performance) over the past five years, can we reveal the main trends in the US economy and make reasonable predictions about future economic trends? The research will use R to help us create and analyze data charts and perform calculations based on our assumptions. Our assumption is that the US economy has experienced a negative state over the past five years, and we will understand the impact of major events on the economy.

## Introduction
What is the purpose of the analysis? What is the question that you are trying to answer with this analysis? Who may this analysis benefit? Are you trying to train a model, test a hypothesis, classify objects apart? Why?

Answer Here:
Analysis Objective: This analysis aims to understand the development of the U.S. economy over the past five years by examining key financial indicators such as GDP growth, inflation, unemployment rate, interest rates, and stock market performance.

Key Questions: The core question is: What are the main trends and drivers behind the changes in the U.S. economy over the past five years? Can these trends tell us where the U.S. economy might be headed in the near future?

Target Audience: This analysis will benefit economists, policymakers, investors, and students majoring in finance and economics who are interested in understanding the structure and future trajectory of the U.S. economy.

Analysis Goals: This analysis aims to test the hypothesis that recent U.S. economic fluctuations and changes in major events are closely related. Furthermore, this analysis involves forecasting, aiming to predict future economic trends based on observed data patterns.

Rationale: This project aims to provide evidence-based insights into the health and future direction of the U.S. economy by combining historical data analysis with predictive modeling.

## Data 
What is the data available for performing the analysis? Where did you retrieve the data from? What are the variables and number of observations available? What steps where performed to clean the data? Will the data be generated through a randomized simulation?

Answer here:
Datasets and Sources:

s&P 500 stocks - Yahoo!
Economic Recession Dates - Federal Reserve
Unemployment Rate - Federal Reserve

Variables and Observations:

Date (Time Period)
GDP Growth Rate (%)
Unemployment Rate (%)
Federal Funds Rate (%)
Major Stock Index Closing Price $

We checked and removed missing or duplicate entries, and to maintain consistency across datasets, we converted the date format and adjusted all variables to the same time frequency (monthly). Random simulations were not performed; all data comes from publicly available financial databases and represents authentic historical records.


##Visualization
Provide preliminary visualization in the form of histograms, density plots, scatter plots, box-plots, and numerical summaries of the data, depending on the type of analysis. What do these visualizations suggest about the hypothesis, model or simulation you are aiming to perform? If you are interested in including further visualization, describe them here.
```{r}

```

##Analysis
Provide some preliminary plans of what analysis you aim to complete. Mention models, tests, procedures, and some ideas of what expected outcomes you hope to find.

Answer here:
After we found out the data and have a question. we will think an hypothesis, make plot, use the plot to anylsis hypothesis, use result to answer the question. we can find and way to perdict next recession. 


## Contributions 
#Jimmy(data analyzer)
- Created repository 
- Created and wrote the Readme file
- Created and worked on the markdown file for Data_Cleaning01, analyzed, cleaned, and combined the data found in the different datasets 
into a single dataset to analyze in Data_Cleaning01

#Charles(data gatherer)
- Help find the data for Fredr(US Recessions), GSPC(S&P 500 stocks),and created the unemployment rate csv file
- Created the usrec_test2.R to get access to the unemployment rate.csv 
- Helped get the code to get Fredr data and S&P 500 into RStudio

#Guanlin(data reporter)
- Created and worked on the 00.requirements.R
- Created and wrote the final report

