---
title: "usrec_markdown"
author: "Jimmy Ieng"
date: "2025-10-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is the Fredr R file converted to an R Markdown file to make the file more organized

#need to figure out how to access BLS
#figure out timeframe to look at


## installs the packages
```{r cars}
#install package, but rstudio can't locate it
install.packages("fredr", dependencies = TRUE)

#checks where the path is located
# should be in a path called LibPath which starts with " and ends with "
installed.packages()["fredr",]

# helps Rstudio finds the path
#replace path with the path your path
.libPaths("/cloud/lib/x86_64-pc-linux-gnu-library/4.5")

#this is to check if its installed, if installed prints TRUE else FAlSE
"fredr" %in% loadedNamespaces()
```
```{r}
library(fredr)
```


help menu for FREDR packages
```{r pressure, echo=FALSE}
?fredr_set_key
```

## unlocks access to Fred
```{r}
fredr_set_key("c859345561941fbef52080a4712c116b")
```


## Charles - "these are needed for every session"
```{r}
## these are needed for every session
library(fredr)
install.packages("fredr")
library("ggplot2")
install.packages("ggplot2")
#library(blsAPI)
```

## pulls up the fredr data
```{r}
usrec <- fredr(series_id = "USREC")
head(usrec)

```

## filtered out data excluding the dates and recession indicator
```{r}
USrec_date <- usrec[,c("date","value")]
```




# pulls up the S&P 500 and the created GSPC_adj to get just the trading volume and the adjusted closing price
```{r}
install.packages("quantmod")
library(quantmod)
getSymbols("^GSPC")  
GSPC_adj <- GSPC[,5:6]
```












